---
title: "bailie23, ISIPTA 2023 submission"
output: html_document
author: "Bailie & Gong"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(jmuOutlier) # for laplace
library(pracma) # for incomplete gamma function
library(Matching)
library(reshape2)
library(ggplot2)
library(rstan)
library(RColorBrewer)
library(xtable)
options(scipen = 10)

mypalette <- brewer.pal(3, 'Set1')
plot.figures <- F
```


# Laplace mechanism for binary sum

```{r}
upper_ml <- function(x, eps, n){
    #eps/2*exp(eps*(n/2 - abs(x - n/2)))
    dlaplace(x = x, n/2, sqrt(2)/eps)*exp(eps*n/2)
}
  
lower_ml <- function(x, eps, n){
    #eps/2*exp(eps*(n/2 - abs(x - n/2)))
    pmin(dlaplace(x = x, n/2, sqrt(2)/eps),
         eps/2*exp(-eps*n/2))*exp(-eps*n/2)
}


plot_ml <- function(nobs, epsilon, color = 'blue'){
  p_ <- ggplot(data.frame(x = seq(-nobs*2, nobs*3, by = 0.1)), aes(x)) +
  geom_function(fun = lower_ml, args = list(eps = epsilon, n = nobs),
                colour = color) +
  theme_bw() +
  labs(title = 'density bounds for p( t | \u03B8 )',  x = 't', y = 'density',
       caption = paste0('n = ', nobs, ', \u03B5 = ', epsilon))


  p_both <- p_ + geom_function(fun = upper_ml, args = list(eps = epsilon, n = nobs), colour = color) #+ labs(subtitle = 'Upper and lower densities')

  p_lower <- p_ + labs(subtitle = 'Lower density only')
  
  return(list(both=p_both, lower=p_lower))
}
```

## Visualizing the marginal likelihood IoM bounds

A few combinations of $n$ and $\epsilon$ values:

```{r}
#plot_ml(nobs = 10, epsilon = 0.1)
#plot_ml(nobs = 10, epsilon = 0.25)
#plot_ml(nobs = 20, epsilon = 0.05)

plot_ml(nobs = 10, epsilon = 0.1, color = mypalette[2])$both
if (plot.figures){
ggsave('marginal1.pdf', width = 4, height = 3.5, device = cairo_pdf)
}

plot_ml(nobs = 10, epsilon = 0.25, color = mypalette[2])$both
if (plot.figures){
ggsave('marginal2.pdf', width = 4, height = 3.5, device = cairo_pdf)
}
```


# Local DP randomized response

```{r}
lower_rr <- function(t, eps){
    (exp(eps) + 1)^(-t)
}
  
upper_rr <- function(t, eps){
  (exp(eps) + 1)^(-t)*exp(t*eps)
}
```

```{r}
epsilon <- 1
nt = 10
dat2 <- data.frame(t = seq(1, nt, by = 1)) %>%
  mutate(lower = lower_rr(t=t, eps = epsilon)) %>%
  mutate(upper = upper_rr(t=t, eps = epsilon)) %>%
  pivot_longer(!t)
#head(dat2)

p_rr <- ggplot(dat2, aes(x = t, y = value, color = name)) +
  geom_line() + geom_point() +
  theme_bw() + scale_color_manual(values = mypalette[c(2, 2)], guide = 'none') +
  scale_x_continuous(breaks = seq(2,10,2)) +
  labs(title = 'density bounds for p( t | \u03B8 )',  x = '|t|', y = 'density',
       caption = paste0('max |t| = ', nt, ', \u03B5 = ', epsilon))


p_rr
if (plot.figures){
ggsave('marginal_rr.pdf', width = 5, height = 3.5, device = cairo_pdf)
}
```


# Posterior bound

Same model as Gong (2022, JPC).  Bayesian Poisson parameter estimation from a clamped single count protected using a Laplace mechanism. Data likelihood is
$$x \mid \theta \sim Pois(\theta).$$
Prior is
$$\theta \sim Gamma(\alpha, 1)$$
for a variety of $\alpha$ values. The privacy mechanism is
$$s = \text{clamp}(x, A) + e, \quad e\sim Lap(\Delta\epsilon^{-1}),$$
where $\text{clamp}(x, A)$ returns the nearest value in a pre-specified set $A$ to $x$. 

The marginal likelihood is

$$\ell(\theta; s) = p(s\mid \theta) = \int \eta(s \mid x)p(x\mid \theta)dx,$$
and thus the Bayesian posterior $p(\theta \mid s) \propto p(\theta)\ell(\theta; s)$ which, due to clamping, does not admit a simple analytical solution. However, this can be simulated via ABC.

```{r}
clamp <- function(x, lower, upper){
  y = x
  y[y < lower] <- lower
  y[y > upper] <- upper
  return(y)
}

post_abc_clamp <- function(y, nsample = 1e3, a = 1, b = 1, 
                           eps = 1, a0 = 0, a1 = 100){
  # Gamma-Pois DP with clamping, ABC algorithm
  # lower, upper: clamping range
  sens <- a1 - a0
  s_ <- sqrt(2)*sens/eps
  temp_ <- function(nsim){
    l_ <- rgamma(nsim, shape = a, rate = b)
    x_ <- rpois(nsim, l_) %>% clamp(lower = a0, upper = a1)
    probs <- dlaplace(y-x_, mean = 0, sd = s_)/dlaplace(0, mean = 0, sd = s_)
    ind <- which(runif(nsim) < probs)
    return(list(lambda = l_[ind], yield = max(1, length(ind))/nsim))
  }
  tp_ <- temp_(nsample)
  lambda <- tp_$lambda; yield = tp_$yield
  cat('Yield is about', yield)
  if (length(lambda) < nsample){
    lambda <- c(lambda, temp_(ceil(nsample/yield))$lambda)
  }
  return(list(lambda, yield))
}


prior_predictive <- function(nsim, a, b, a0, a1, eps){
    l_ <- rgamma(nsim, shape = a, rate = b)
    x_ <- rpois(nsim, l_) %>% clamp(lower = a0, upper = a1)
    t_ <- x_ + rlaplace(nsim, mean = 0, sd = sqrt(2)*(a1-a0)/eps)
    return(list(x = x_, t = t_))
}
```



We obtain bounds for the posterior $p(\theta \mid s)$ in relation to $p(\theta)$.


```{r}
alpha <- 3; beta <- 1; # hyperparameters
a0 <- 0; a1 <- 6
epsilon <- 1

set.seed(10)
nruns = 10
pp0 <- prior_predictive(nruns, a = alpha, b = beta, a0 = a0, a1 = a1, eps = epsilon)

set.seed(100)
draws = list()
for (i in 1:nruns){
  run_ <- post_abc_clamp(y = pp0$t[i], nsample = 2e3, 
                         a = alpha, b = beta, eps = epsilon, a0 = a0, a1 = a1)
  draws[[i]] <- run_[[1]]
}


draws.dat <- data.frame(sapply(draws, function(x) x[1:max(lengths(draws))])) %>%
  pivot_longer(everything())

ll <- seq(range(draws.dat$value, na.rm = T)[1], range(draws.dat$value, na.rm = T)[2], by = 0.05)

dat <- data.frame(lambda = ll) %>%
  mutate(prior = dgamma(lambda, shape = alpha, rate = beta)) %>%
  mutate(upper = prior*exp(epsilon)) %>%
  mutate(lower = prior*exp(-epsilon)) %>%
  pivot_longer(!lambda)
#head(dat)

p <- ggplot(dat, aes(x = lambda, y = value, linetype = name)) +
  geom_density(data = draws.dat, aes(group = name, x = value), color = 'gray80', inherit.aes = F) +
  geom_line(color = mypalette[2]) +
  scale_linetype_manual(values = c('solid', 'dashed', 'solid'), guide = 'none') +
  theme_bw() +
  labs(x = '\u03B8', y = 'posterior density', title = 'density bounds for posterior p( \u03B8 | t )',
      caption = paste0('prior for \u03B8 is Gamma(', alpha, ',', beta, '), clamping range is [', a0, ',', a1, '], \u03B5 = ', epsilon))

p
if (plot.figures){
ggsave('posterior.pdf', width = 5, height = 3.5, device = cairo_pdf)
}
```
